Distributional Trees and Forests

### 2017-10-13 pkg/disttree/inst/draft/draft.Rnw
further changes and new text in section probabilistic precipitation forecasting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1660 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c74d4c49c115f5b487d790e5b1b5703ba5dfc235

\Abstract{
In the field of probabilistic forecasting classical linear models such as ensemle model output statistics (EMOS) are still common and widely used. This is due to their simplicity and their interpretability. However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility. Generalized additive models for location, scale, and shape (GAMLSS) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM. A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models. However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods. Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from the meteorology or based on a computationally more demanding boosting approach.

@@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage
\vspace{0.0cm}
\minipage{0.25\textwidth}
\begin{center}
\hspace{0.4cm}
LM, GLM 
\end{center}
\endminipage
\hspace{1.4cm}
\minipage{0.25\textwidth}
\begin{center}
GAM
\end{center}
\endminipage
\hspace{1.4cm}
\minipage{0.25\textwidth}
\begin{center}
GAMLSS
\end{center}
\endminipage
\caption{\label{Devel_parmod}Development in the field of parametric modeling}
\end{figure}

@@

The exact way of determining how and where to split varies over different tree algorithms. For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm. Summed up, the steps of building a distributional tree using one of these algorithms are
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\bold{\hat{\theta}}$ via maximum likelihood.
\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose the breakpoint with the highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is met.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}
The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.\\
Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.\\
\\
For distributional trees the strategy of how to make a prediction for a new observation $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward. All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new observation are selected and a distributional model is fit on the subset. To select this subset a weight vector $\bold{\omega^{tree}} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
$$
\omega^{tree}_i(\bold{z}^{\ast}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z}^{\ast} \in \mathcal{B}_b))
$$
The predicted distribution of the new observation is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
$$
\hat{\theta}(\bold{z}^{\ast}) = \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{tree}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i).
$$

@@

\SweaveOpts{eval = TRUE}
\newpage
\section{Rain}
In this application we analyze precipitation data meassured over 28 years at 95 (?) stations all over Tyrol and Vorarlberg. The considered data is provided by ..... and consists of the meassured total precipiation within 24 hours of each day in July from 1985 until 2012, both included. Additionally, numerical forecasts from the ECMWF are used as covariates. These forecasts are ensemble predictions consisting of ... individual forecasts based on perturbed initiall conditions.\\
In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max. All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standarad deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
The following table shows all variables together with the number of variations and the provided variations listed in the last column.\\
\\
The response variable total precicipitation is power transformed by a factor of 1.6. This is done due to experience and knowledge about possible distributions of precipitation meassurements. The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.

@@

\newpage
\begin{table}[h]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{| l  c  l |}
\hline
Basic covariates & {\#} & Variations\\
\hline
\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sum, \\
\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sum, \\ 
\textbf{cape}: convective                  &    & ensemble minimum of sum, \\
\hspace*{0.4cm} available potential energy &    & ensemble maximum of sum\\
\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all over 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sum over\\
\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std dev of sum over\\
\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean values, \\
\hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of minimal values,\\
\textbf{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\textbf{pwat}: preciptable water          &   & ensemble std dev of mean values,\\
\textbf{tmax}: 2m maximum                 &   & ensemble std dev of minimal values,\\
\hspace*{1.15cm} temperature              &   & ensemble std dev of maximal values,\\
\textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{1.05cm} condensate  &   & \\
\textbf{t500}: temperature   &   & \\
\hspace*{1cm} on 500 hPA     &   & \\
\textbf{t700}: temperature   &   & \\
\hspace*{1cm} on 700 hPA     &   & \\
\textbf{t850}: temperature   &   & \\
\hspace{1cm} on 850 hPA      &   & \\
\hline
\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPA &   & ensemble minimum of difference in mean,\\
\textbf{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPA &   & \qquad all over 6--30 UTC\\
\textbf{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPA &   & \\
\hline
\textbf{msl{\_}diff}: mean sea level  & 1 variation: & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
\hspace{1.65cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number and the type of variations}
\label{covariates}
\end{minipage} 
%}
\end{table}

\newpage

The following methods are applied on this data set:

\begin{table}[h]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{| l | l   l   l |}
\hline
 & \textbf{Type} & \textbf{Location ($\mu$)} & \textbf{Scale ($\sigma$)} \\
\hline
\textbf{Forest} & recursive    & all & all \\ 
                & partitioning &     &     \\ 
\hline
\textbf{GAM (prespecified)} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
                            & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
 & & tp{\_}mean1218 $\cdot$ & tp{\_}sprd1218 $\cdot$\\ 
 & & \quad cape{\_}mean1218, & \quad  cape{\_}mean1218,\\
 & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
 & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
 & & pwat{\_}mean{\_}mean, & \\
 & & tdiff500850{\_}mean, & \\
 & & msl{\_}diff & \\
\hline
\textbf{GAM (boosted)} & spline  & all & all \\
                       & in each &     &     \\
\hline
\textbf{EMOS} & linear & tp{\_}mean & tp{\_}spread \\
\hline
\end{tabular}
\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter}
\label{model_specification}
\end{minipage} 
%}
\end{table}

For the gamlss model the covariates were selected manually based on expert knowledge. Additionally, interactions are specified in advance, once between tppow{\_}mean1218 and capepow{\_}mean1218 for the location parameter and once between tppow{\_}sprd1218 and capepow{\_}mean1218 for the l ocatioin parameter. By putting extra effort in the specification of the formulas the gamlss model is provided with additional information and is therefore expected to profit from this advantage.\\

>>>>

\Abstract{
In the field of probabilistic forecasting classical linear models such as ensemle model output statistics (EMOS) are still common and widely used. This is due to their simplicity and their interpretability. However, the restriction to modelling only linear effects requires further extensions which allow for more flexibility. Generalized additive models for location, scale, and shape (GAMLSS, \cite{stasinopoulos2005}) can also capture non-linear effects as each distribution parameter can be modelled separately by its own GAM. A different approach that would make it possible to also model non-additive effects and interactions is to apply tree structured models. However, up to now, this class of models has not yet been integrated in probabilistic forecasting as no general distributional framework is available for these methods. Therefore, we are going to introduce distributional trees and forests which embed the idea of flexible parametric modeling into the framework of regression trees and random forests. Applying these novel approaches on precipitation data in a mountainous region with a large number of numerical weather prediction quantities illustrates their strengths as their performance is compared to GAMLSS models that are specified based either on subject-matter knowledge/experience from the meteorology or based on a computationally more demanding boosting approach.

@@

<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x, y, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x, rev(x)), c(ytrue + 0.1 + var, rev(ytrue - 0.1 - var)),
  col = pallight["forest"], border = "transparent")
lines(x, ytrue, col = pal["forest"], lwd=7)
points(x, y, col = "slategray", pch = 19)
box(lwd = 5)
@
\end{center}
\endminipage
\vspace{0.0cm}
\minipage{0.25\textwidth}
\begin{center}
\hspace{0.4cm}
LM, GLM 
\end{center}
\endminipage
\hspace{1.7cm}
\minipage{0.25\textwidth}
\begin{center}
GAM
\end{center}
\endminipage
\hspace{1.7cm}
\minipage{0.25\textwidth}
\begin{center}
GAMLSS
\end{center}
\endminipage
\caption{\label{Devel_parmod}Development in the field of parametric modeling}
\end{figure}

@@

The exact way of determining how and where to split varies over different tree algorithms. (For distributional trees, one of two algorithms can be chosen as a framework to build the tree: either the MOB algorithm or the ctree algorithm.) Summed up, the steps of building a distributional tree using one of these algorithms are
\begin{enumerate}
\item Specify a distribution with log-likelihood function $\ell(\theta; y)$.
\item Estimate $\bold{\hat{\theta}}$ via maximum likelihood.
\item Test for associations or instabilities of the scores $s(\bold{\hat{\theta}}, y_i)$ and each partitioning variable $Z_l$.
%\item Assess whether the \emph{model scores} are associated with
%    (or change along) any of the available covariates -- e.g.,
%    using parameter instability tests (\emph{strucchange}) or
%    conditional inference (\emph{coin}).
\item Split the sample along the partitioning variable with the strongest association or instability.
    Choose the breakpoint with the highest improvement in log-likelihood.
\item Repeat steps 2--4 recursively in the subgroups
    until some stopping criterion is met.
    %-- e.g., for significance or sample size.
%\item Choose the variable with the strongest association
%\item Choose the split point which leads to the highest improvement of the model
%\item Split and repeat 2-6 in each node until a stopping criterion is met
\end{enumerate}
The stopping criterion in step 5 can for example be a minimum number of observations in a node or a p-value for the statistical tests applied in step 3.\\
Executing these steps results in a stepwise parametric model where a complete distribution is fit in each node of the tree.\\
\\
For distributional trees the strategy of how to make a prediction for a new observation $\bold{z}^{\ast} = (z_1^{\ast}, \ldots, z_m^{\ast})$ is very simple and straight forward. All observations of the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ which end up in the same terminal node of the tree as the new observation are selected and a distributional model is fit on the subset. To select this subset a weight vector $\bold{\omega^{tree}} = (\omega^{tree}_1, \ldots, \omega^{tree}_n)$ is included in the fitting process.
$$
\omega^{tree}_i(\bold{z}^{\ast}) = \sum_{b=1}^B I((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z}^{\ast} \in \mathcal{B}_b))
$$
The predicted distribution of the new observation is then fully specified by the estimated parameter $\hat{\theta}(\bold{z}^{\ast})$ where
$$
\hat{\theta}(\bold{z}^{\ast}) = \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{tree}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i).
$$

@@

\SweaveOpts{eval = TRUE}
\newpage
\section{Probabilistic Forecasting of Precipitation}
As explained in the introduction distributional trees and forests provide a new approach in probabilistic forecasting which combines advantages of parametric models that are already used in this field and tree structured models which provide features that could be particularly valueable in probabilistic forecasting but have not yet been applied in this kind of settings. To show their strengths in practice, these new methods are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities. In particular, the data set consists of daily 24h precipitation sums for the month of July from 28 years, starting in 1985 until 2012, both included. The observations were meassured at 95 (?) stations all over Tyrol (and regions close to the frontier). The considered data is provided by ..... 
Next to the variable of interest, the meassured total precipiation within 24 hours of each day, numerical forecasts from the ECMWF are used as covariates. These forecasts are ensemble predictions consisting of ... individual forecasts based on perturbed initiall conditions.\\
In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions, noted as ensemble mean, ensemble std dev, ensemble min and ensemble max. All of these values can be calculated on the whole day (6--30 UTC) or a 6h time window of the day.\\
Another possible option is to consider the mean, minimum or maximum over 24 hours for each individual prediction and then take the ensemble mean or ensemble standarad deviation, noted as ensemble mean of mean/min/max or ensemble std dev of mean/min/max respectively.\\
Table \ref{covariates} shows all variables together with the number of variations and the provided variations listed in the last column.\\
The response variable total precicipitation is power transformed by a factor of 1.6. This is done due to experience and knowledge about possible distributions of precipitation meassurements. The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.

@@

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{| l  c  l |}
\hline
Basic covariates & {\#} & Variations\\
\hline
\textbf{tp}: total precipitation            & 12 variations:  & ensemble mean of sum, \\
\hspace*{\fill} power transformed (by 1.6) &    & ensemble std dev of sum, \\ 
\textbf{cape}: convective                  &    & ensemble minimum of sum, \\
\hspace*{0.4cm} available potential energy &    & ensemble maximum of sum\\
\hspace*{\fill}  power transformed (by 1.6)&    & \qquad all over 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sum over\\
\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std dev of sum over\\
\hspace*{\fill}                            &    & \qquad 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\textbf{dswrf}: downwards short ware      & 6 variations: & ensemble mean of mean values, \\
\hspace*{\fill} radiation flux (sunshine) &   & ensemble mean of minimal values,\\
\textbf{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\textbf{pwat}: preciptable water          &   & ensemble std dev of mean values,\\
\textbf{tmax}: 2m maximum                 &   & ensemble std dev of minimal values,\\
\hspace*{1.15cm} temperature              &   & ensemble std dev of maximal values,\\
\textbf{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{1.05cm} condensate  &   & \\
\textbf{t500}: temperature   &   & \\
\hspace*{1cm} on 500 hPA     &   & \\
\textbf{t700}: temperature   &   & \\
\hspace*{1cm} on 700 hPA     &   & \\
\textbf{t850}: temperature   &   & \\
\hspace{1cm} on 850 hPA      &   & \\
\hline
\textbf{tdiff500850}: temperature         & 3 variations: & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPA &   & ensemble minimum of difference in mean,\\
\textbf{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPA &   & \qquad all over 6--30 UTC\\
\textbf{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPA &   & \\
\hline
\textbf{msl{\_}diff}: mean sea level  & 1 variation: & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
\hspace{1.65cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number and the type of variations}
\label{covariates}
\end{minipage} 
%}
\end{table}

Next to distributional forests other methods / the same methods as in the simulation section are applied to this data set and their results are compared based on RMSE, log-likelihood values and CRPS values. The chosen methods and their specifications are listed in Table \ref{model_specification}.
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{| l | l   l   l |}
\hline
 & \textbf{Type} & \textbf{Location ($\mu$)} & \textbf{Scale ($\sigma$)} \\
\hline
\textbf{Forest} & recursive    & all & all \\ 
                & partitioning &     &     \\ 
\hline
\textbf{GAM (prespecified)} & spline  & tp{\_}mean, & tppow{\_}sprd,\\
                            & in each & tp{\_}max,  & dswrf{\_}sprd{\_}mean,\\
 & & tp{\_}mean1218 $\cdot$ & tp{\_}sprd1218 $\cdot$\\ 
 & & \quad cape{\_}mean1218, & \quad  cape{\_}mean1218,\\
 & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
 & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
 & & pwat{\_}mean{\_}mean, & \\
 & & tdiff500850{\_}mean, & \\
 & & msl{\_}diff & \\
\hline
\textbf{GAM (boosted)} & spline  & all & all \\
                       & in each &     &     \\
\hline
\textbf{EMOS} & linear & tp{\_}mean & tp{\_}spread \\
\hline
\end{tabular}
\caption[Table caption text]{Table of applied methods together with their type (way of including the covariates) and the included covariates for each distribution parameter}
\label{model_specification}
\end{minipage} 
%}
\end{table}

For the gamlss model the covariates were selected manually based on meteorological knowledge and experience. Additionally, interactions are specified in advance, once between tp{\_}mean1218 and cape{\_}mean1218 for the location parameter and once between tp{\_}sprd1218 and cape{\_}mean1218 for the locatioin parameter. By putting extra effort in the specification of the formulas the gamlss model is provided with additional information and is therefore expected to profit from this advantage.\\
